\chapter{Lightweight Deep Learning Refinement}
\label{ch:dl}

Layer~3 of the \projectshort{} architecture (\chapref{ch:system}) adds an optional deep-learning stage \emph{after} scene-aware mathematical denoising (\chapref{ch:routing}) and \emph{before} evaluation or ASR ingestion. This chapter specifies the TinyResidualRefiner student network, the DeepFilterNet3-based distillation pipeline, training and hyperparameter selection, benchmark results, and deployment characteristics. The design goal is to transfer a fraction of teacher quality into a sub-10\,KB checkpoint that runs in milliseconds on CPU, without replacing the interpretable math backbone at inference.

\section{Design Philosophy: Math-First, DL-Assisted}
\label{sec:dl_philosophy}

\subsection{Motivation}

Large end-to-end enhancers such as DeepFilterNet3~\cite{deepfilternet2023} achieve strong perceptual quality but require millions of parameters, non-trivial memory, and GPU-class throughput for comfortable real-time use (\chapref{ch:background}, Table~\ref{tab:method_tradeoffs}). For local ASR preprocessing---where latency, battery, and explainability matter---deploying a full teacher on every utterance is often impractical.

\projectshort{} therefore adopts a \emph{hybrid} strategy:
\begin{enumerate}
  \item \textbf{Math-first backbone} --- routed OM-LSA/MCRA and specialized front-ends provide a fast, inspectable baseline $D_{\mathrm{math}}$ (\chapref{ch:algorithms}).
  \item \textbf{Residual DL correction} --- a tiny student $\mathcal{F}_\theta$ predicts only a bounded multiplicative mask on the STFT of $D_{\mathrm{math}}$, not clean speech from scratch.
  \item \textbf{Offline teacher} --- DeepFilterNet3 generates training targets; it is \emph{not} invoked at inference when refinement is disabled or when only the student is deployed.
\end{enumerate}

This division implements the research gap identified in \chapref{ch:background}: preserve modular interpretability and scene routing while closing part of the quality gap to a strong neural teacher.

\subsection{Residual-on-Math Formulation}

Let $Y$ denote the noisy STFT and $D_{\mathrm{math}}=\mathcal{G}_{\mathrm{route}}(Y)$ the complex STFT after mathematical enhancement. The student refines
\begin{equation}
  \hat{Y} = M \odot D_{\mathrm{math}}, \qquad
  M = \mathrm{clip}\bigl(1 + \alpha\,\tanh(\mathcal{F}_\theta(\mathbf{X})),\; m_{\min},\, m_{\max}\bigr),
  \label{eq:residual_mask}
\end{equation}
with $m_{\min}=0.2$, $m_{\max}=1.8$, and residual scale $\alpha=0.8$ (training and default inference). Phase is inherited from $D_{\mathrm{math}}$:
\begin{equation}
  \angle \hat{Y}(k,t) = \angle D_{\mathrm{math}}(k,t).
\end{equation}
Magnitude-domain residual learning reduces the dynamic range the network must represent compared to mapping $Y\mapsto \hat{S}$ directly~\cite{hintondistill2015}, while keeping enhancement decisions partially auditable through the routed baseline.

\begin{remark}[Why not distill end-to-end?]
An end-to-end student would need to internalize routing, WPE, notch banks, and OM-LSA inside one black box. By distilling on \emph{routed} inputs, the student specializes in \emph{closing residual errors} left by the current math pipeline---consistent with Theorem~\ref{thm:chain} and the analysis--design loop (\chapref{ch:routing}).
\end{remark}

\section{End-to-End Distillation Pipeline}
\label{sec:distill_pipeline}

\paragraph{Stage 1: Routed math outputs.}
\texttt{run\_pipeline.py} produces \texttt{*\_routed.wav} for each benchmark scene using \texttt{choose\_route()} and override config (\chapref{ch:routing}).

\paragraph{Stage 2: Teacher generation.}
\texttt{build\_distill\_pairs.py} invokes the DeepFilterNet CLI (\texttt{deepFilter}) with weights under \texttt{models/DeepFilterNet3/}, writing \texttt{*\_teacher.wav} per scene.

\paragraph{Stage 3: Pair manifest.}
Each entry in \texttt{distill\_pairs.json} links $(y_{\mathrm{noisy}}, D_{\mathrm{math}}, T_{\mathrm{teacher}}, s_{\mathrm{clean}})$ with aligned file paths. An 80/20 scene split (\texttt{distill\_split.json}) holds out scenes 13--15 for validation.

\paragraph{Stage 4: Student training.}
\texttt{train\_student\_distill.py} optimizes $\theta$ with mask-distillation and compressed-magnitude losses for up to 1200 steps or 900\,s.

\paragraph{Stage 5: Refinement inference.}
\texttt{infer\_student\_refine.py} or the web \texttt{run\_distill\_refine} flag applies the checkpoint (\texttt{student\_runD.pt}) to routed audio.

\section{TinyResidualRefiner Architecture}
\label{sec:architecture_student}

\subsection{Input Representation}

Given routed waveform $d_{\mathrm{math}}(n)$ and noisy $y(n)$, compute STFTs $D_{\mathrm{math}}(k,t)$ and $Y(k,t)$ with $N_{\mathrm{FFT}}=512$, hop $R=128$, Hann window (matching \texttt{common.py} and OM-LSA). Log-compressed magnitudes form a two-channel map:
\begin{equation}
  \mathbf{X}(k,t) =
  \begin{bmatrix}
    \log(1+|D_{\mathrm{math}}(k,t)|) \\
    \log(1+|Y(k,t)|)
  \end{bmatrix}
  \in \mathbb{R}^{2\times F\times T},
  \label{eq:student_input}
\end{equation}
where $F=257$ one-sided bins at 16\,kHz. Log1p compression stabilizes gradients across 40\,dB dynamic range while preserving low-energy consonant structure.

\subsection{Convolutional Stack}

The student $\mathcal{F}_\theta$ is a three-layer Conv2D network (\texttt{TinyResidualRefiner} in \texttt{train\_student\_distill.py}):
\begin{align}
  \mathbf{H}_1 &= \mathrm{ReLU}\bigl(\mathrm{Conv}_{3\times3}^{2\to 16}(\mathbf{X})\bigr), \\
  \mathbf{H}_2 &= \mathrm{ReLU}\bigl(\mathrm{Conv}_{3\times3}^{16\to 16}(\mathbf{H}_1)\bigr), \\
  \mathbf{r}   &= \tanh\bigl(\mathrm{Conv}_{1\times1}^{16\to 1}(\mathbf{H}_2)\bigr) \in [-1,1]^{F\times T}.
\end{align}
Each frequency--time location receives a $3\times3$ local context in $\mathbf{X}$; the $1\times1$ head produces a scalar residual per bin. The predicted mask follows \eqref{eq:residual_mask} with $\alpha$ stored in the checkpoint for inference scaling.

\begin{table}[H]
  \centering
  \caption{TinyResidualRefiner layer shapes and parameter counts.}
  \label{tab:student_params}
  \begin{tabular}{lcccc}
    \toprule
    \textbf{Layer} & \textbf{Kernel} & \textbf{In$\to$Out} & \textbf{Params} & \textbf{Output shape} \\
    \midrule
    Conv2d + ReLU   & $3\times3$ & $2\to16$  & 304  & $16\times F\times T$ \\
    Conv2d + ReLU   & $3\times3$ & $16\to16$ & 2320 & $16\times F\times T$ \\
    Conv2d + Tanh   & $1\times1$ & $16\to1$  & 17   & $F\times T$ \\
    \midrule
    \textbf{Total}  &            &           & \textbf{2641} & \\
    \bottomrule
  \end{tabular}
\end{table}

\subsection{Design Rationale}

\begin{itemize}
  \item \textbf{Two-channel input} --- $|D_{\mathrm{math}}|$ exposes what the math chain already removed; $|Y|$ preserves noise context for scenes where routing under-suppresses.
  \item \textbf{Tanh residual} --- bounds raw network output before clipping, preventing explosive masks early in training.
  \item \textbf{Symmetric clip $[0.2,1.8]$} --- allows both attenuation and slight boost relative to routed magnitude, matching teacher mask dynamic range.
  \item \textbf{No recurrent layers} --- entire utterance processed in one STFT pass; latency scales linearly with length and is dominated by STFT/iSTFT, not Conv2D.
\end{itemize}

\section{Training Data Construction}
\label{sec:distill_pairs}

\subsection{Pair Schema}

For each scene $s$ in \texttt{noisy\_testset/}, \texttt{build\_distill\_pairs.py} emits:
\begin{center}
\small
\begin{tabular}{ll}
  \texttt{noisy\_wav}      & original mixture $y$ \\
  \texttt{math\_input\_wav} & routed output $d_{\mathrm{math}}$ (\texttt{*\_routed.wav}) \\
  \texttt{teacher\_wav}    & DeepFilterNet3 output on $y$ \\
  \texttt{clean\_wav}      & \texttt{clean\_reference.wav} (shared) \\
\end{tabular}
\end{center}

Teacher inference uses the bundled DeepFilterNet3 checkpoint; outputs are cached under \texttt{distill/teacher/} unless \texttt{--regen-teacher} is set.

\subsection{Train/Validation Split}

With 15 paired scenes, the split file fixes:
\begin{itemize}
  \item \textbf{Train (12 scenes):} scene01--scene12;
  \item \textbf{Validation (3 scenes):} scene13 (\texttt{street\_combo}), scene14 (\texttt{office\_fan}), scene15 (\texttt{reverb\_plus\_hiss}).
\end{itemize}
Validation scenes were not seen during gradient updates but share the same noise taxonomy as training, providing a conservative generalization check before AB evaluation on all 15 outputs.

\section{Knowledge Distillation Training}
\label{sec:distill_training}

\subsection{Teacher-Derived Mask Target}

Let $X$, $T$, $C$ be complex STFTs of routed, teacher, and clean waveforms on a random 1.0\,s chunk. Teacher mask targets are
\begin{equation}
  M_{\mathrm{teacher}}(k,t) =
  \mathrm{clip}\!\left(
    \frac{|T(k,t)|}{|X(k,t)|+\varepsilon},\;
    m_{\min},\, m_{\max}
  \right).
  \label{eq:teacher_mask}
\end{equation}
This ratio expresses how much additional suppression (or boost) the teacher applies relative to the math baseline---a direct soft target for mask distillation rather than raw waveform matching.

\subsection{Compressed-Magnitude Loss}

To align refined STFT magnitudes with teacher and clean references, define the compressed magnitude with exponent $c=0.5$:
\begin{equation}
  \psi_c(Z) = \bigl(|Z|+\varepsilon\bigr)^{c}.
\end{equation}
The refinement $\hat{Y}=M\odot X$ minimizes
\begin{equation}
  \mathcal{L} =
    \lambda_1 \,\| M - M_{\mathrm{teacher}} \|_1
  + \lambda_2 \,\| \psi_c(\hat{Y}) - \psi_c(T) \|_2^2
  + \lambda_3 \,\| \psi_c(\hat{Y}) - \psi_c(C) \|_2^2,
  \label{eq:distill_loss}
\end{equation}
with defaults $(\lambda_1,\lambda_2,\lambda_3)=(1.0, 0.3, 0.15)$ from \texttt{distill\_config.json}. The L1 mask term transfers teacher suppression patterns; compressed-magnitude terms stabilize training in low-SNR bins where linear MSE on power over-penalizes silence.

\subsection{Optimization Details}

Training uses Adam with learning rate $5\times10^{-4}$ (selected runD uses $3.5\times10^{-4}$ via \texttt{--lr-scale 0.7}), gradient clipping at norm 1.0, and one random 1.0\,s chunk per step ($16\,000$ samples at 16\,kHz). Loop termination occurs at 1200 steps or 900\,s elapsed---whichever comes first. Checkpoints store \texttt{model\_state}, STFT sizes, \texttt{residual\_alpha}, and validation L1 mask error.

\begin{algorithm}[H]
  \caption{Student distillation training step (\texttt{train\_student\_distill.py})}
  \label{alg:distill_train}
  \begin{algorithmic}[1]
    \Require Train pairs $\mathcal{P}$, config, TinyResidualRefiner $\mathcal{F}_\theta$
    \Ensure Checkpoint \texttt{student\_\{tag\}.pt}
    \State Build sampling probabilities from \texttt{hard\_scene\_weights}
    \For{$\mathit{step} = 1$ \textbf{to} \texttt{max\_steps} (or time limit)}
      \State Sample pair $(X_{\mathrm{audio}}, T_{\mathrm{audio}}, Y_{\mathrm{audio}}, C_{\mathrm{audio}})$ by scene weight
      \State Extract aligned 1.0\,s chunks; compute STFTs $X,T,Y,C$
      \State $M_{\mathrm{teacher}} \gets \mathrm{clip}(|T|/(|X|+\varepsilon),\, 0.2,\, 1.8)$ \Comment{\eqref{eq:teacher_mask}}
      \State $\mathbf{X} \gets [\log(1+|X|);\, \log(1+|Y|)]$
      \State $r \gets \mathcal{F}_\theta(\mathbf{X})$;\quad $M \gets \mathrm{clip}(1+\alpha r,\, 0.2,\, 1.8)$
      \State $\hat{Y} \gets M \odot X$
      \State $\mathcal{L} \gets \lambda_1\|M-M_{\mathrm{teacher}}\|_1 + \lambda_2\|\psi_c(\hat{Y})-\psi_c(T)\|_2^2 + \lambda_3\|\psi_c(\hat{Y})-\psi_c(C)\|_2^2$
      \State Adam step; clip gradients
    \EndFor
    \State Evaluate val L1 on held-out scenes; save checkpoint
  \end{algorithmic}
\end{algorithm}

\section{Hard-Scene Weighted Sampling}
\label{sec:hard_scenes}

Uniform scene sampling under-represents difficult benchmarks where math routing leaves large teacher--math gaps. Let $w_s$ be the weight for scene $s$. Training draws pair index $i$ with probability
\begin{equation}
  P(i) = \frac{w_{s(i)}}{\sum_j w_{s(j)}}.
\end{equation}
Default weights in \texttt{distill\_config.json}:
\begin{center}
\begin{tabular}{lc}
  \texttt{scene02\_white\_snr5} & $3.0$ \\
  \texttt{scene09\_chirp\_interference\_snr8} & $3.5$ \\
  \texttt{scene11\_intermittent\_bursts\_snr7} & $1.5$ \\
  all other train scenes & $1.0$ \\
\end{tabular}
\end{center}

Scene02 ($\rho_{<0}\approx 0.90$, \chapref{ch:scene}) and scene09 (chirp ridge; largest routed RMS error in eval) receive the highest weights, aligning optimization with scenes where residual correction yields the largest $\Delta$RMS (Table~\ref{tab:distill_eval}).

\section{Training Configuration and Model Selection}
\label{sec:train_config}

\subsection{Hyperparameters}

Table~\ref{tab:distill_config} lists defaults. STFT parameters match the math library so mask learning operates in the same time--frequency tiling as OM-LSA.

\begin{table}[H]
  \centering
  \caption{Default distillation hyperparameters (\texttt{distill\_config.json}).}
  \label{tab:distill_config}
  \begin{tabular}{lc}
    \toprule
    \textbf{Hyperparameter} & \textbf{Value} \\
    \midrule
    Sample rate             & 16\,kHz \\
    STFT ($N_{\mathrm{FFT}}$, hop) & 512, 128 \\
    Chunk length            & 1.0\,s (16\,000 samples) \\
    Max training time       & 900\,s \\
    Max steps               & 1200 \\
    Base learning rate      & $5 \times 10^{-4}$ \\
    Residual scale $\alpha$ & 0.8 \\
    Compress exponent $c$  & 0.5 \\
    $(\lambda_1, \lambda_2, \lambda_3)$ & (1.0, 0.3, 0.15) \\
    Selected checkpoint tag & \texttt{runD} \\
    \bottomrule
  \end{tabular}
\end{table}

\subsection{Hyperparameter Search and \texttt{runD} Selection}

\texttt{hparam\_search\_report.json} documents a two-phase search:
\begin{enumerate}
  \item \textbf{Phase 1} --- reference run with conservative inference residual scale;
  \item \textbf{Phase 2} --- compare candidates \texttt{runC} ($\mathrm{lr}=5\times10^{-4}$) and \texttt{runD} ($\mathrm{lr}=3.5\times10^{-4}$) by validation mask L1, then AB RMS on the full benchmark.
\end{enumerate}
\texttt{runD} achieved $\mathrm{val\_L1}\approx 0.522$ with 1200 steps in 20.1\,s on CPU (\texttt{train\_log\_runD.json}) and was selected because it improved routed RMS in all 15 scenes at inference scale 1.0 (mean $\Delta$RMS $=6.19\times10^{-4}$).

\section{Benchmark Evaluation}
\label{sec:distill_eval}

\texttt{eval\_distill\_ab.py} compares three waveforms per scene against \texttt{clean\_reference.wav}:
\begin{equation}
  e_{\mathrm{routed}} = \mathrm{RMS}(d_{\mathrm{math}}-s),\quad
  e_{\mathrm{refined}} = \mathrm{RMS}(\hat{s}-s),\quad
  e_{\mathrm{teacher}} = \mathrm{RMS}(t-s).
\end{equation}

Table~\ref{tab:distill_eval} summarizes \texttt{distill\_eval.json}. The student improves over routed output in \textbf{all 15 scenes} (mean $\Delta$RMS $=6.19\times10^{-4}$). Largest gains occur on structured-interference scenes aligned with hard-scene weighting: scene09 (chirp, $\Delta=0.00191$) and scene12 (babble, $\Delta=0.00143$). Scene04 is the only case where the student beats the teacher on RMS ($e_{\mathrm{refined}}<e_{\mathrm{teacher}}$), indicating teacher--clean mismatch on colored noise rather than student superiority in general.

\begin{table}[H]
  \centering
  \caption{Distillation A/B evaluation (\texttt{student\_runD.pt}, RMS error vs.\ clean).}
  \label{tab:distill_eval}
  \footnotesize
  \begin{tabular}{lrrrr}
    \toprule
    \textbf{Scene} & $e_{\mathrm{routed}}$ & $e_{\mathrm{refined}}$ & $e_{\mathrm{teacher}}$ & $\Delta$ \\
    \midrule
    scene01 white 15\,dB  & 0.00623 & 0.00587 & 0.00441 & 0.00036 \\
    scene02 white 5\,dB   & 0.01003 & 0.00936 & 0.00766 & 0.00067 \\
    scene03 pink          & 0.00808 & 0.00755 & 0.00529 & 0.00053 \\
    scene04 brown         & 0.00555 & 0.00527 & 0.01076 & 0.00028 \\
    scene05 hum50         & 0.00538 & 0.00476 & 0.00340 & 0.00062 \\
    scene06 hum60         & 0.00554 & 0.00512 & 0.00347 & 0.00043 \\
    scene07 HF hiss       & 0.00655 & 0.00617 & 0.00406 & 0.00038 \\
    scene08 LF rumble     & 0.00673 & 0.00631 & 0.00435 & 0.00042 \\
    scene09 chirp         & 0.01686 & 0.01495 & 0.00447 & 0.00191 \\
    scene10 clicks        & 0.00845 & 0.00802 & 0.00626 & 0.00043 \\
    scene11 bursts        & 0.01412 & 0.01373 & 0.00655 & 0.00039 \\
    scene12 babble        & 0.01379 & 0.01236 & 0.01080 & 0.00143 \\
    scene13 street        & 0.00932 & 0.00868 & 0.00727 & 0.00064 \\
    scene14 office fan    & 0.00785 & 0.00737 & 0.00578 & 0.00048 \\
    scene15 reverb+hiss   & 0.00724 & 0.00691 & 0.00476 & 0.00033 \\
    \midrule
    \textbf{Mean}         &       &         &         & 0.00062 \\
    \bottomrule
  \end{tabular}
\end{table}

\begin{proposition}[Monotonic refinement on benchmark (empirical)]
\label{prop:distill_improve}
Under the fixed \texttt{runD} checkpoint and inference $\alpha=1.0$, distilled refinement satisfies $e_{\mathrm{refined}} \le e_{\mathrm{routed}}$ for all 15 benchmark scenes in \texttt{distill\_eval.json}. Mean improvement is modest in absolute RMS ($\sim\!6\times10^{-4}$) but consistent, supporting optional refinement without degrading any scene relative to routed math output.
\end{proposition}

The student remains above the teacher on most scenes ($e_{\mathrm{refined}} > e_{\mathrm{teacher}}$): distillation transfers \emph{partial} teacher knowledge into 2641 parameters, not full DeepFilterNet capacity.

\section{Inference and Deployment}
\label{sec:inference}

\subsection{Offline Batch Inference}

\texttt{infer\_student\_refine.py} loads \texttt{student\_runD.pt}, reads each \texttt{math\_input\_wav} from \texttt{distill\_pairs.json}, and writes \texttt{*\_distill\_refined.wav} to \texttt{outputs/distill\_refined/}. Optional \texttt{--residual-scale} multiplies checkpoint $\alpha$ for conservative refinement (phase~1 search used scale 0.1).

\begin{algorithm}[H]
  \caption{Student refinement inference (\texttt{infer\_student\_refine.py})}
  \label{alg:distill_infer}
  \begin{algorithmic}[1]
    \Require Routed WAV $d_{\mathrm{math}}$, noisy WAV $y$, checkpoint $\theta$
    \Ensure Refined WAV $\hat{s}$
    \State $D_{\mathrm{math}} \gets \mathrm{STFT}(d_{\mathrm{math}})$;\quad $Y \gets \mathrm{STFT}(y)$
    \State $\mathbf{X} \gets [\log(1+|D_{\mathrm{math}}|);\, \log(1+|Y|)]$
    \State $M \gets \mathrm{clip}(1 + \alpha\,\mathcal{F}_\theta(\mathbf{X}),\, 0.2,\, 1.8)$
    \State $\hat{Y} \gets M \odot D_{\mathrm{math}}$
    \State $\hat{s} \gets \mathrm{iSTFT}(\hat{Y})$;\quad peak-normalize if $|\hat{s}|_\infty > 0.99$
    \State \Return $\hat{s}$
  \end{algorithmic}
\end{algorithm}

\subsection{Web Platform Integration}

When the API request sets \texttt{run\_distill\_refine=true} (\texttt{schemas/models.py}), \texttt{distill\_refiner.refine\_with\_student()} runs after routing:
\begin{enumerate}
  \item Mathematical denoising produces intermediate \texttt{processed/} WAV;
  \item Student loads \texttt{distill/checkpoints/student\_runD.pt};
  \item Refinement uses routed output as $D_{\mathrm{math}}$ and original upload as $Y$;
  \item Refined audio replaces or supplements the processed artifact for metrics/plots.
\end{enumerate}
If the checkpoint is missing, refinement is skipped gracefully (\texttt{return False}).

\subsection{Deployment Artifact Size}

The \texttt{student\_runD.pt} checkpoint contains 2641 float32 weights ($\approx 10$\,KB) plus metadata---orders of magnitude smaller than DeepFilterNet3 model directories. Combined with the math library (no GPU requirement), this supports edge deployment as an optional post-router stage.

\section{Computational Advantage}
\label{sec:compute_adv}

Table~\ref{tab:compute_compare} contrasts teacher and student on this project's 16\,kHz benchmark setting. Parameter counts are exact for TinyResidualRefiner; DeepFilterNet3 is reported at order-of-millions per published architectures~\cite{deepfilternet2023} and bundled \texttt{config.ini} (256-dim embedding, multi-layer GRU/DF stacks).

\begin{table}[H]
  \centering
  \caption{Computational comparison: DeepFilterNet3 teacher vs.\ TinyResidualRefiner student.}
  \label{tab:compute_compare}
  \begin{tabular}{lcc}
    \toprule
    \textbf{Metric} & \textbf{DeepFilterNet3} & \textbf{TinyResidualRefiner} \\
    \midrule
    Parameters        & $\sim\!10^6$ (teacher) & 2\,641 \\
    Checkpoint size   & $\sim$MB scale        & $\sim$10\,KB \\
    Used at inference & Optional baseline only & Optional refine flag \\
    Training time     & Pretrained (external) & $\approx 20$\,s / 1200 steps (CPU) \\
    Latency (5.6\,s clip, CPU) & $\approx 2.9$\,s CLI wall time & $\approx 8$\,ms STFT+Conv+iSTFT \\
    Speed ratio       & 1$\times$            & $\sim\!350\times$ faster (refine only) \\
    \bottomrule
  \end{tabular}
\end{table}

\paragraph{Interpretation.}
Measured student latency on \texttt{scene01} ($5.6$\,s audio) is dominated by STFT/iSTFT ($\approx 8$\,ms total on CPU), with Conv2D forward pass under 1\,ms for $F\times T\approx 257\times 440$. DeepFilterNet3 CLI processing on the same file requires $\approx 2.9$\,s wall time including model load and 48\,kHz-native inference---unsuitable as a per-utterance default on low-power devices. The hybrid pipeline keeps math routing ($\sim$ tens of ms) as the primary cost and adds $<$10\,ms optional refinement.

\paragraph{Quality--cost trade-off.}
Student RMS error sits between routed math and teacher on most scenes (Table~\ref{tab:distill_eval}): users gain consistent incremental improvement at negligible marginal latency, while the teacher remains available offline for re-distillation when routing or math modules change.

\section{Limitations and Extensions}
\label{sec:dl_limits}

\begin{itemize}
  \item \textbf{Routed-input dependence} --- The student corrects errors of the \emph{current} router and overrides; changing \texttt{scene\_method\_overrides.json} without retraining may misalign mask statistics.
  \item \textbf{Magnitude-only mask} --- Phase is copied from $D_{\mathrm{math}}$; scenarios where phase errors dominate (heavy reverberation) may need complex-domain extensions.
  \item \textbf{Teacher domain gap} --- DeepFilterNet3 defaults to 48\,kHz training; teacher targets on 16\,kHz benchmark clips introduce domain mismatch (evident on scene04 where teacher RMS exceeds routed).
  \item \textbf{Single-module overrides vs.\ chains} --- Distillation pairs use whatever routed WAV exists; multi-stage chains (hum: notch$\rightarrow$NMF) are folded into one $D_{\mathrm{math}}$ without stage-wise students.
  \item \textbf{RMS-only selection} --- \texttt{runD} was chosen by RMS AB metrics; perceptual or ASR-word-error criteria may rank checkpoints differently.
\end{itemize}

Future work includes on-device INT8 quantization of $\mathcal{F}_\theta$, joint retraining after router updates, and word-error-rate evaluation on downstream ASR (\chapref{ch:experiments}).

\section{Chapter Summary}
\label{sec:dl_summary}

This chapter presented the lightweight deep-learning layer of \projectshort{}: a 2641-parameter TinyResidualRefiner that applies a bounded residual mask to routed STFT magnitudes, trained by distilling DeepFilterNet3 mask and compressed-magnitude targets with hard-scene weighted sampling. Training completes in about 20 seconds on CPU; inference adds roughly 8\,ms per 5.6-second utterance. On all 15 benchmark scenes, \texttt{student\_runD.pt} improves RMS error versus routed mathematical output while remaining interpretable as a correction on top of logged algorithm chains. \chapref{ch:web} describes web UI exposure; \chapref{ch:experiments} places distillation results in the full experimental study.
