\chapter{Conclusion and Future Work}
\label{ch:conclusion}

\section{Summary}
\label{sec:summary}

This report presented \projectshort{} (\projectname{}), an intelligent audio denoising analysis platform aimed at \emph{local preprocessing} for speech-centric applications. The system integrates three layers: (1)~data-driven scene characterization and ANASS parameter guidance; (2)~an enhanced OM-LSA/MCRA library with seven composable physics-matched modules and adaptive routing; and (3)~an optional TinyResidualRefiner trained by distillation from DeepFilterNet3. A web workflow supports upload, batch processing, diagnostic visualization, and informal perceptual testing.

\subsection{Answers to Research Questions}

\textbf{RQ1 (Scene adaptivity).}
Full-dimensional analysis of 15 noise scenarios (\chapref{ch:scene}) demonstrated that global SNR labels alone are insufficient: band-wise imbalance, local-SNR tails, and Cohen's $d$ separability differ sharply across scenes. Grid-search override routing---selecting the best single module per scene from scene-informed candidates---reduced mean RMS error from $0.00936$ to $0.00837$ (10.6\% relative improvement), beating monolithic OM-LSA on 13/15 scenes. Module winners (\texttt{nmf\_denoise} on five scenes, \texttt{subspace\_denoise} on four, \texttt{wpe\_omlsa} on three) align with the interference taxonomy used to build the library. Scene03 confirmed that adaptive selection should include a \emph{no-op} fallback to OM-LSA when noise is near-stationary. Live rule routing plus override files implements the same design philosophy with auditable \texttt{reason} strings (\chapref{ch:routing}).

\textbf{RQ2 (Algorithm design).}
The enhanced OM-LSA/MCRA core replaces fixed spectral subtraction with MMSE gains, MCRA noise tracking, and speech-presence soft gating (\chapref{ch:algorithms}). Mixed chains (WPE, notch, subspace, Kalman, NMF, chirp cancellation followed by OM-LSA) address interference classes that violate single-estimator assumptions. On structured scenes, specialized routing yields large gains---e.g., scene10 clicks RMS $0.01062 \to 0.00844$ ($-20.5\%$) with NMF, and isolated \texttt{chirp\_notch} on scene09 reaching $0.01185$ versus base $0.01506$. Diagnostic fixed subtraction in scene reports correctly predicted high musical-noise risk (kurtosis 142 on clicks).

\textbf{RQ3 (Hybrid refinement).}
TinyResidualRefiner runK (56\,129 parameters, $\alpha=0.8$) improved routed RMS on \textbf{all 15 scenes}, lowering the mean from $0.00837$ to $0.00730$ and closing approximately 44\% of the mean gap to the DeepFilterNet3 teacher ($0.00596$) while adding only $\approx 8$\,ms refine latency on CPU. Gains were largest on chirp and bursts ($\Delta e = 0.00436$ and $0.00569$), where math routing leaves structured residual error. Mean SNR improved from 12.69\,dB (routed) to 14.06\,dB (runK). No scene regressed relative to its routed WAV under \texttt{student\_runK.pt}.

\textbf{RQ4 (Evaluability).}
The web platform (\chapref{ch:web}) provides REST APIs, 28 Plotly metrics per task, residual export, history management, and ABX sessions. Experiments in \chapref{ch:experiments} were reproduced from JSON artifacts (\texttt{distill\_eval.json}, \texttt{best\_method\_search.json}, \texttt{routed\_route\_log.json}) and script-generated spectrograms. Docker packaging and pinned dependencies support repeatable demos for coursework and external review.

\subsection{Main Quantitative Outcomes}

Table~\ref{tab:final_outcomes} summarizes the enhancement cascade on the shared 15-scene benchmark (5.618\,s, 16\,kHz mono, aligned to \texttt{clean\_reference.wav}).

\begin{table}[H]
  \centering
  \caption{Final benchmark outcomes (mean RMS error vs.\ clean reference).}
  \label{tab:final_outcomes}
  \begin{tabular}{lcc}
    \toprule
    \textbf{Pipeline stage} & \textbf{Mean $e_{\mathrm{RMS}}$} & \textbf{Scenes improved vs.\ previous stage} \\
    \midrule
    Monolithic OM-LSA/MCRA & 0.00936 & --- \\
    Per-scene override routing & 0.00837 & 13 / 15 \\
    + TinyResidualRefiner (\texttt{runK}) & 0.00730 & 15 / 15 \\
    DeepFilterNet3 teacher (offline) & 0.00596 & 14 / 15 vs.\ runK \\
    \bottomrule
  \end{tabular}
\end{table}

Qualitative spectrograms on low-SNR white, chirp, babble, and reverb+hiss scenes confirm that routing removes the dominant interference mechanism and refinement sharpens speech-dominated structure without introducing gross distortion at the scales displayed.

\section{Future Work}
\label{sec:future}

The following directions extend the current prototype toward production-grade ASR front-ends and stronger scientific claims.

\begin{enumerate}
  \item \textbf{Learned scene router.}
  Replace hand-tuned priority rules with a classifier or small MLP on the scene feature vector ($\mathcal{M}(s)$ in \chapref{ch:routing}), trained to predict module or chain labels from grid-search oracle targets. This would reduce reliance on \texttt{scene\_method\_overrides.json} while preserving log interpretability if the model outputs calibrated probabilities and top-$k$ rationales.

  \item \textbf{Online ANASS parameter adaptation.}
  Integrate Module~1--3 noise estimators and over-subtraction controls (\chapref{ch:scene}) into the streaming path so the first 200--500\,ms of an utterance updates $\hat{N}(k,t)$ and gain floors before the bulk of speech is processed---critical for mobile ASR where scene JSON is unavailable.

  \item \textbf{Richer distillation.}
  Retrain runK on aug-expanded pairs (\texttt{augment\_data.py}); experiment with multi-teacher ensembles and phase-aware masks. The merged repository supports 9500+ aug entries for large-scale retraining when augmented WAVs are available.

  \item \textbf{Real-time streaming and mobile export.}
  Chunk STFT processing with overlap-add state buffers; export TinyResidualRefiner to ONNX or TFLite and benchmark on ARM devices. Latency targets should be stated jointly with WER or MOS, not RMS alone.

  \item \textbf{Expanded corpora and formal listening tests.}
  Add recorded real-world noise (café, vehicle, office), more talkers, and languages. Run double-blind ABX and MOS with $\ge 20$ trials per condition and multiple listeners~\cite{loizou2013}, using the web UI in blind mode.

  \item \textbf{Downstream ASR coupling.}
  Wire enhanced audio into a fixed Mandarin ASR engine and report word error rate with and without preprocessing---the ultimate metric motivated in \chapref{ch:introduction}. Compare \projectshort{} against RNNoise and full DeepFilterNet under identical decoder settings.

  \item \textbf{Pipeline hygiene.}
  Automate regeneration of \texttt{*\_routed.wav} when overrides change; add CI checks that \texttt{distill\_eval.json} matches on-disk WAV hashes to prevent scene09-style log--file drift.
\end{enumerate}

\section{Closing Remarks}
\label{sec:closing}

From a machine-learning course perspective, \projectshort{} demonstrates how classical signal processing and modern deep learning can be composed rather than opposed: scene statistics supply supervised \emph{structure} for routing and training curricula; MMSE estimators provide a stable, interpretable backbone; and knowledge distillation transfers representational power from a large teacher into a deployable student without shipping the teacher to end users. Students and practitioners can trace every benchmark output from JSON metrics through route logs to WAV artifacts and spectrograms---an educational workflow that single end-to-end black-box enhancers rarely afford.

For speech-enhancement practice, the project shows that heterogeneous real-world noise is not well served by one algorithm or one neural model. The 15-scene benchmark is small but deliberately diverse; the consistent refinement gains and routing ablation suggest that \emph{selecting the right physics first} and \emph{learning small residuals second} is a viable engineering pattern for edge ASR preprocessing. The open web platform turns that pattern into an reproducible demo suitable for further experimentation.

In summary, \projectshort{} delivers a documented, evaluable hybrid denoising stack with measurable RMS improvements, transparent routing, and sub-second CPU inference for typical chains---while clearly marking the remaining gap to state-of-the-art neural teachers and the need for ASR- and perceptually grounded evaluation in future iterations. The analysis--design--validation loop embodied in this report is intended to outlive any single checkpoint or router version: as new scenes are added, the same toolchain can re-analyze, re-route, re-distill, and re-quantify without rewriting the core platform architecture.
